Aligning marine species range data to better serve science and conservation

### 2016-03-21 data_explore/data_explore_global_maps.Rmd
incorporated Ben's comments; reframing a conclusion; preparing figures.  Moved the MPA analysis explanation of shifts into the SI

@@

```{r global map matched species, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

cat('First off: map-paired species.  Map 1: AM dataset; map 2: IUCN dataset; maps 3 & 4: log-transformed AM & IUCN\n')

### find total species in each dataset for later percent calcs
# am_n_tot   <- am_spp_cells %>% 
#   filter(am_sid %in% unique(spp_list$am_sid)) %>%
#   dplyr::select(am_sid) %>% 
#   unique() %>% nrow()
# iucn_n_tot <- iucn_spp_cells %>% 
#   filter(sciname %in% unique(spp_list$sciname)) %>%
#   dplyr::select(sciname) %>% 
#   unique() %>% nrow()
### good - the same number

### Number of aquamaps species per cell
iucn_pair_rast_file <- file.path(dir_data, 'rasters/iucn_pair_rast.tif')
am_pair_rast_file   <- file.path(dir_data, 'rasters/am_pair_rast.tif')

if(!file.exists(iucn_pair_rast_file) | !file.exists(am_pair_rast_file)) {
  am_n_per_cell <- am_spp_cells %>%
  filter(am_sid %in% unique(spp_list$am_sid)) %>%
  group_by(loiczid) %>%
  summarize(count_am = n()) %>%
#   summarize(count_am = n(),
#             pct_am = count_am/am_n_tot) %>%
  as.data.frame()

  am_sp_map <- subs(loiczid_raster, am_n_per_cell[ , c('loiczid', 'count_am')], 
                    by = 'loiczid', 
                    which = 'count_am', 
                    subsWithNA = TRUE)

  ### Number of IUCN species per cell
  iucn_n_per_cell <- iucn_spp_cells %>%
    group_by(loiczid) %>%
    summarize(count_iucn = n()) %>%
  #   summarize(count_iucn = n(),
  #             pct_iucn = count_iucn/iucn_n_tot) %>%
    as.data.frame()

  iucn_sp_map <- subs(loiczid_raster, iucn_n_per_cell[ , c('loiczid', 'count_iucn')], 
                      by = 'loiczid', 
                      which = 'count_iucn', 
                      subsWithNA = TRUE)

  writeRaster(iucn_sp_map, filename = iucn_pair_rast_file, overwrite = TRUE)
  writeRaster(am_sp_map, filename = am_pair_rast_file, overwrite = TRUE)
} else {
  am_sp_map   <- raster(am_pair_rast_file)
  iucn_sp_map <- raster(iucn_pair_rast_file)
}

am_ln_map <- log(am_sp_map + 1)
iucn_ln_map <- log(iucn_sp_map + 1)

### Plot both maps
myTheme <- rasterTheme(region = brewer.pal('Blues', n = 9))

levelplot(am_sp_map, par.settings = myTheme, 
#          main = 'Aquamaps Dataset', 
          colorkey = list( at = seq(0, 1063, length = 15)))

levelplot(iucn_sp_map, par.settings = myTheme, 
#          main = 'IUCN Dataset', 
          colorkey = list(at = seq(0, 1063, length = 15)))

levelplot(am_ln_map, par.settings = myTheme, 
#          main = 'Aquamaps Dataset', 
          colorkey = list( at = seq(0, maxValue(am_ln_map), length = 15)))

levelplot(iucn_ln_map, par.settings = myTheme, 
#          main = 'IUCN Dataset', 
          colorkey = list(at = seq(0, maxValue(iucn_ln_map), length = 15)))

Now look at full aquamaps species

```{r all_aquamaps, eval = FALSE, echo = FALSE}

cat('Full AM dataset, species richness\n')

@@

  ### find total spp to allow for percent calcs later
  # am_all_n_tot <- am_spp_cells_big %>% 
  #   dplyr::select(SpeciesID) %>% 
  #   unique() %>% nrow()

  all_am <- am_spp_cells_big %>%
    group_by(loiczid) %>%
    summarize(all_count = n()) %>%
  #   summarize(all_count = n(),
  #             all_pct   = all_count/am_all_n_tot) %>%
    as.data.frame()

@@

am_levelplot1 <- levelplot(all_am_map, 
#                           main = 'Entire Aquamaps Dataset',
                           par.settings = myTheme)

# all_am_ln_map <- log(all_am_map + 1)
# all_am_ln_map <- all_am_ln_map/maxValue(all_am_ln_map)
# am_levelplot1 <- levelplot(all_am_ln_map, par.settings = myTheme, main = 'Entire Aquamaps Dataset: log-transformed relative species richness', 
#           colorkey = list(at = seq(0, 1, length = 15)))

print(am_levelplot1)
# print(am_levelplot2)

# am_ln_pts <- as.data.frame(rasterToPoints(all_am_ln_map))
# spp_richness_am_plot <- ggplot(am_ln_pts, aes(x = x, y = y)) +
#   ggtheme_map + 
#   geom_raster(aes(fill = layer), alpha = .7) +
#   scale_fill_continuous(low = 'white', high = 'blue') +
#   borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
#   scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
#   scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
#   labs(title = 'AquaMaps Dataset: log-transformed relative species richness', x = NULL, y = NULL, fill = NULL) 
# 
# spp_richness_am_plot
# ggsave(file.path(dir_fig, 'all_am_ln_map.png'), height = 4, width = 8)

@@

cat('Full IUCN dataset, species richness\n')

@@

  ### find total spp to allow for percent calcs later
  # am_all_n_tot <- am_spp_cells_big %>% 
  #   dplyr::select(SpeciesID) %>% 
  #   unique() %>% nrow()

  all_iucn <- iucn_spp_cells_all %>%
    group_by(loiczid) %>%
    summarize(all_count = n()) %>%
  #   summarize(all_count = n(),
  #             all_pct   = all_count/am_all_n_tot) %>%
    as.data.frame()

@@

} else {
  all_iucn_map   <- raster(all_iucn_rast_file)
}

iucn_levelplot1 <- levelplot(all_iucn_map, 
                             #main = 'Entire IUCN Dataset',
                             par.settings = myTheme)

# all_iucn_ln_map <- log(all_iucn_map + 1)
# all_iucn_ln_map <- all_iucn_ln_map/maxValue(all_iucn_ln_map)
# iucn_levelplot2 <- levelplot(all_iucn_ln_map, 
#           par.settings = myTheme, 
#           margin = list(axis = TRUE),
#           main = 'Entire IUCN Dataset: log-transformed relative species richness', 
#           colorkey = list(at = seq(0, 1, length = 15)))

print(iucn_levelplot1)
# print(iucn_levelplot2)
# iucn_ln_pts <- as.data.frame(rasterToPoints(all_iucn_ln_map))
# spp_richness_iucn_plot <- ggplot(iucn_ln_pts, aes(x = x, y = y)) +
#   ggtheme_map + 
#   geom_raster(aes(fill = layer), alpha = .7) +
#   scale_fill_continuous(low = 'white', high = 'blue') +
#   borders('world', color='gray40', fill='gray45', size = .1) +  # create a layer of borders
#   scale_x_continuous(breaks = seq(-180, 180, by = 30), expand = c(0, 2)) +
#   scale_y_continuous(breaks = seq( -90,  90, by = 30), expand = c(0, 2)) +
#   labs(title = 'IUCN Dataset: log-transformed relative species richness', x = NULL, y = NULL, fill = NULL) 
# 
# spp_richness_iucn_plot
# ggsave(file.path(dir_fig, 'all_iucn_ln_map.png'), height = 4, width = 8)

>>>>

```{r setup, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

# Libraries and Paths

library(ggplot2)
library(maps)
library(readr)
library(data.table)
library(raster)
library(tidyr)
library(dplyr)
library(stringr)

dir_N <- c('Windows' = '//neptune.nceas.ucsb.edu/data_edit',
           'Darwin'  = '/Volumes/data_edit',
           'Linux'   = '/var/data/ohi')[[ Sys.info()[['sysname']] ]]

dir_git <- '~/github/IUCN-AquaMaps'

dir_anx <- file.path(dir_N, 'git-annex/globalprep/SPP_ICO')
dir_fig <- file.path(dir_git, 'figures')
dir_data <- file.path(dir_git, 'data')

if(basename(getwd()) != 'data_explore') setwd('data_explore')
source('data_explore_fxns.R')
### When knitting this, it automatically sets WD to be this directory...
### the 'setwd()' is there for running by chunk
```

@@

```{r global_maps, echo = FALSE, warning = FALSE, message = FALSE, eval = FALSE}

@@

myTheme <- rasterTheme(region = brewer.pal('Blues', n = 9))

message('Full AM dataset, species count\n')

@@

  all_am <- am_spp_cells_big %>%
    group_by(loiczid) %>%
    summarize(all_count = n()) %>%
    as.data.frame()

@@

### set up boundaries - from http://oscarperpinan.github.io/rastervis/FAQ.html#sec-7
ext <- as.vector(extent(all_am_map))

boundaries <- maps::map('world', fill=TRUE,
    xlim=ext[1:2], ylim=ext[3:4],
    plot=FALSE)

## read the map2SpatialPolygons help page for details
IDs <- sapply(strsplit(boundaries$names, ":"), function(x) x[1])
bPols <- maptools::map2SpatialPolygons(boundaries, IDs = IDs,
                              proj4string = CRS(projection(all_am_map)))

### create plot, add boundary polygons
am_levelplot1 <- levelplot(all_am_map, 
                           scales = list(at = seq(-180, 180, 45)),
                           xlab = NULL, ylab = NULL,
                           par.settings = myTheme) + 
  layer(sp.polygons(bPols, fill = 'grey90', col = 'grey30', lwd = .25))

png(file.path(dir_fig, 'all_aquamaps_map1.png'),
    width = 8.7, height = 6, units = 'in', res = 300)
print(am_levelplot1)
dev.off()

@@

message('Full IUCN dataset, species count\n')

@@

  all_iucn <- iucn_spp_cells_all %>%
    group_by(loiczid) %>%
    summarize(all_count = n()) %>%
    as.data.frame()

@@

} else {
  all_iucn_map   <- raster(all_iucn_rast_file)
  values(all_iucn_map)[is.na(values(all_am_map))] <- NA
}

iucn_levelplot1 <- levelplot(all_iucn_map, 
                             scales = list(at = seq(-180, 180, 45)),
                             xlab = NULL, ylab = NULL,
                             par.settings = myTheme) + 
  layer(sp.polygons(bPols, fill = 'grey90', col = 'grey30', lwd = .25))

png(file.path(dir_fig, 'all_iucn_map1.png'),
    width = 8.7, height = 6, units = 'in', res = 300)
print(iucn_levelplot1)
dev.off()

### 2016-03-21 ms_1_abst_sig.Rmd
incorporated Ben's comments; reframing a conclusion; preparing figures.  Moved the MPA analysis explanation of shifts into the SI

Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and show inconsistencies.  AquaMaps (Kaschner et al. ___2015???___) and the International Union for Conservation of Nature (IUCN) (IUCN 2015) are two distinct efforts to map marine species distributions at a global scale.  Together these two databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` species included in both.  Here we examine differences in predicted species ranges between the two datasets, and find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors.  We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index (Halpern et al. 2012, 2015) and a global analysis of gaps in coverage of marine protected areas (Klein et al. 2015) - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on how the two datasets were prioritized. Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making will be essential.

@@

(223 words; 1593 char with spaces)

>>>>

Species distribution data provide the foundation for a wide range of ecological research studies and conservation management decisions, yet most species ranges remain unknown, and existing range maps often suffer from data limitations and show inconsistencies.  AquaMaps and the International Union for Conservation of Nature (IUCN) are two distinct efforts to map marine species distributions at a global scale.  Together these databases represent `r format(n_spp_all, big.mark = ',')` species (`r format(p_spp_am, big.mark = ',')`% within AquaMaps, `r format(p_spp_iucn, big.mark = ',')`% within IUCN), with only `r format(n_spp_both, big.mark = ',')` shared species.  Here we examine differences in predicted species ranges between the two datasets and find that these misalignments mainly result from divergent methodologies that introduce differing frequencies of commission and omission errors.  We illustrate the scientific and management implications of these differences by repeating two recent applications - an assessment of global biodiversity within the Ocean Health Index and a global analysis of gaps in coverage of marine protected areas - and find significantly different results depending on how the two datasets were used.  Until a single, highly accurate dataset of global marine species ranges becomes available, understanding the implications of dataset differences for conservation planning and decision-making is essential.

@@

(184 words; 1307 char with spaces)

### 2016-03-21 ms_2_intro_res_disc.Rmd
incorporated Ben's comments; reframing a conclusion; preparing figures.  Moved the MPA analysis explanation of shifts into the SI

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps (Kaschner et al. ___2013???___) and International Union for Conservation of Nature (IUCN) (IUCN 2015).  These two spatial datasets have been used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, 2015 _both_; Selig et al. 2013 _both_), evaluating global biodiversity patterns (Coll et al. 2010 _AM_; Martin et al. 2014 _both and others_, Pimm et al. 2014 _IUCN, mostly focusing on terrestrial but some marine_, Kaschner et al. 2011), predicting range shifts (Molinos et al. 2015 _AquaMaps_), and setting conservation priorities (Klein et al. 2015 _AquaMaps_).

___BSH:___ _I think Julie makes this comment below, but we may need to be able to substantiate this somehow.  Is there a way to quickly figure out how many studies have used AM vs. IUCN?_

___CCO:___ _What about age - AquaMaps is pretty recent (2006 at earliest); before that, would IUCN have been the dominant dataset for marine?  Maybe we can track citations of AquaMaps, but citations of IUCN are generic - just cite the Red List - not referencing spatial information specifically, much less marine.  Going by citation year field on the maps themselves, the oldest marine record I can find is 2004._

_Three foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; Ready et al. 2010) have been cited a total of 315 times._

@@

We then reexamine two recent marine biodiversity studies - an assessment of the status of global biodiversity within the Ocean Health Index (Halpern et al. 2012, 2015) and a global analysis of gaps in protection afforded by marine protected areas (MPAs) (Klein et al. 2015) - as case studies to explore the implications of prioritizing one data set over the other.  The results highlight the possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

@@

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" (IUCN 2015) - essentially a refined extent of occurrence, based on observation records and refined by expert understanding of the species' range and habitat preferences.  In contrast, AquaMaps models species distribution based environmental preferences (e.g. temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global raster of "probability of occurrence" for each species (Kaschner et al. 2006, Ready et al. 2010).  Studies frequently define "presence" for AquaMaps by applying a probability threshold, e.g. 40% or greater probability of occurrence (Halpern et al. 2012, 2015).

@@

The two datasets have notably different taxonomic (Fig. 1) and regional (Fig. 2) coverage.  AquaMaps encompasses a broader range of taxa than IUCN, as IUCN spatial data files are only available for select taxonomic groups that have been comprehensively assessed. While species numbers in both datasets peak in tropical latitudes near the equator, species counts for IUCN maps drops quickly beyond 30°N and 30°S, while species counts for AquaMaps remains robust well into temperate latitudes.  The longitude frequency plots show a slight bias in the IUCN dataset away from the Atlantic and eastern Pacific compared to AquaMaps.  To achieve more comprehensive global coverage of species ranges these two datasets can be used together, but the underlying methodological differences complicate such direct comparisons.

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e. where on the map) against the area alignment (ratio of smaller range area to larger range area, i.e. how much of the map) for each shared species (Fig. 3A). 
 This analysis revealed a general negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

@@

The upper right quadrant includes species (n = `r n_q1`) whose described ranges are above average in both spatial distribution and extent. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 3B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because their ranges span nearly the entire map.

@@

While we examined IUCN Red List extinction risk categories by quadrant, expecting to see more concurrence between maps for species of concern; however, we found no compelling patterns (SI Fig. XXX).

@@

We recalculated the OHI species status score under several scenarios to observe the impact of toggling the prioritized data set from IUCN to AquaMaps, and toggling the AquaMaps presence threshold from 40% to 0% (<a href = '#fig4'>Fig. 4</a>).  
 Reducing the threshold increases the apparent range of a species; the slight decrease in average score for scenario 1 suggests increased spatial representation of threatened species.  
 The slight bump in mean score for scenario 2 may indicate a small increase in spatial representation of low-risk species, a small decrease in spatial representation of high-risk species, or a combination of both. 
 The large decrease shown in scenario 3 indicates that a zero threshold greatly increases the spatial representation of high-risk species.

Given that only `r format(p_both_all, big.mark = ',')` percent of species overlap, it is surprising that changing the priority for overlapping species from IUCN maps to AquaMaps would result in such large country-level score shifts as seen in scenarios 2 and 3.  While many country scores showed only small variation across all three scenarios, select countries gained up to 5 points while others dropped as many as 12 points.  Significantly, the global score decreased by two full points in scenario 3, indicating that an arbitrary change in how the two datasets are combined can result in a more dire assessment of species conservation status.

@@

Klein et al. (2015) compare the global distribution of species to the global distribution of marine protected areas to assess how well the MPAs protect key species and identify which species fall through gaps in protection.  The study relied on the AquaMaps database, using a probability of occurrence threshold of 50% or greater, to determine species presence, and the World Database of Protected Areas to define zones of marine protection.  The study determined that the global MPA network is woefully inadequate in protecting marine biodiversity: 90.5% of species have less than 5% of their overall range represented within MPAs and 1.4% of species have no protection at all ("gap" species).

 (<a href = '#fig5'>Figure 5</a>) displays the results of applying this methodology under new scenarios.  When comparing the conclusions based upon an IUCN-only scenario to an AquaMaps-only scenario (using 2015 AquaMaps data and a 0% threshold sets the most meaningful comparison) we found a five-fold increase in the proportion of gap species (6.4% of species vs. 1.2%) and dramatically larger proportion of species with less than 2% of their range protected (73.2% of species vs. 47.7%).  However, this comparison also indicates a larger proportion of well-protected species with greater than 10% of range protected (2.9% of species vs. 1.5%).  Conclusions drawn from each of these datasets would paint dramatically different pictures of the protection afforded by our current global MPA network.

___If you have room, I think you could add a bit more interpretation of these results here. Could discuss the commission and omission stuff here too to describe these somewhat non-intuitive results??? CS___

- check which species are gap species and look at some maps
- check which species are 10%+ species and look at some maps
- where do these species fall georegionally? is there anything odd about their ranges?

@@

AquaMaps and IUCN range maps provide important spatial information on marine species ranges, and show reasonable agreement for many well-studied species, but illustrate uncertainty in our understanding of spatial distribution for many others.  Identifying and addressing differences in these datasets will increase their utility for ecology research and conservation actions.  We have identified several likely drivers of commission errors and omission errors between these datasets, pointing toward improvements to better align the predictions of species distributions.

The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (IUCN 2015).  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; data users might consider a simple cutoff at the photosynthetic limit of 200 meters to minimize a likely source of commission errors.

The AquaMaps methodology is a powerful means of quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently, these area restrictions are denoted by UN Food and Agriculture Organisation (FAO) fisheries reporting areas (Ready et al. 2010); however, area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (Spalding et al. 2007), would likely decrease commission errors and improve predictive power, especially for data-poor species.

When using AquaMaps data alone, changing the probability of occurrence threshold to determine presence becomes a tradeoff between errors of commission (low threshold) and omission (high threshold).  Using AquaMaps in conjunction with IUCN can increase the taxonomic and spatial breadth of coverage, as long as the differences between the datasets can be reasonably minimized. In this case, we recommend a presence threshold of 0% as it most closely approximates the "limits of distribution" criterion defined by IUCN data providers.

<!-- * Additional filters on AquaMaps distribution criteria, such as expert review status or the number of occurrence cells used to generate the species model, can help avoid overextrapolation for data poor species. _using other attributes and variables to filter? this seems outside the scope of this paper_ -->

_(do we need a wrap-it-up paragraph here?  I put that more at the top of the conclusions section... could shift it down here, but probably not needed in both places)_

>>>>

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps (Kaschner et al. ___2013???___) and International Union for Conservation of Nature (IUCN) (IUCN 2015).  These two spatial datasets have been used in hundreds of studies and applications for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, 2015; Selig et al. 2013), evaluating global biodiversity patterns (Coll et al. 2010; Martin et al. 2014, Pimm et al. 2014, Kaschner et al. 2011), predicting range shifts (Molinos et al. 2015), and setting conservation priorities (Klein et al. 2015).

<!--
Three foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; Ready et al. 2010) have been cited a total of 315 times.
-->

@@

We then reexamine two recent marine biodiversity studies - an assessment of the status of global biodiversity within the Ocean Health Index (Halpern et al. 2012, 2015) and a global analysis of gaps in protection afforded by marine protected areas (MPAs) (Klein et al. 2015) - as case studies to explore the implications of prioritizing one data set over the other.  The results highlight possible consequences of different data use decisions on our understanding of marine biodiversity status and protection.

@@

The IUCN publishes species range maps based on expert input of spatial boundaries of a given species' "limits of distribution" (IUCN 2015) - essentially a refined extent of occurrence, based on observation records and informed by expert understanding of species' range and habitat preferences.  In contrast, AquaMaps models species distribution based on environmental preferences (e.g., temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global raster of "probability of occurrence" for each species (Kaschner et al. 2006, Ready et al. 2010).  Studies frequently define "presence" for AquaMaps by applying a probability threshold, e.g., the Ocean Health Index defines presence as 40% or greater probability of occurrence (Halpern et al. 2012, 2015).

@@

The two datasets have notably different taxonomic (Fig. 1) and regional (Fig. 2) coverage.  AquaMaps encompasses a broader range of taxa than IUCN, as IUCN spatial data files are only available for select taxonomic groups that have been comprehensively assessed. While species numbers in both datasets peak in tropical latitudes near the equator, species counts for IUCN maps drop quickly beyond 30°N and 30°S, while species counts for AquaMaps remain robust well into temperate latitudes.  The longitude frequency plots show a slight bias in the IUCN dataset away from the Atlantic and eastern Pacific compared to AquaMaps.  To achieve more comprehensive global coverage of species ranges these two datasets can be used together, but the underlying methodological differences complicate such direct comparisons.

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (how much of the smaller range falls within the larger range, i.e., where on the map) against the area alignment (ratio of smaller range area to larger range area, i.e., how much of the map) for each shared species (Fig. 3A). 
 This analysis revealed a general negative linear pattern, suggesting that increasing similarity in range area correlates with decreasing distribution alignment.
 AquaMaps tends to extrapolate species ranges into suitable areas beyond known occurrences, such that each additional unit of range predicted by AquaMaps will fall in different locations than an additional unit of range predicted using IUCN methodology. For species with dissimilar range areas, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar range areas, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

@@

The upper right quadrant includes species (n = `r n_q1`) whose described ranges are above average in both spatial distribution and extent. These species tend to be well-studied and include wide-ranging pelagic organisms such as marine mammals, tunas, and billfishes (Fig. 3B). This result is not surprising, as species with very large ranges are likely to be more aligned regardless of methodology simply because their ranges span nearly the entire map.  We expected to see significant concurrence between maps for IUCN Red List species of concern, assuming that increased attention would lead to improved understanding of species ranges; however, we found no compelling patterns to indicate that this was the case (SI Fig. XXX).

@@

@@

We recalculated the OHI species status score under several scenarios to observe the impact of toggling the prioritized data set from IUCN to AquaMaps, and toggling the AquaMaps presence threshold from 40% to 0% (<a href = '#fig4'>Fig. 4</a>).  Reducing the threshold increases the apparent range of a species; the slight decrease in average score for scenario 1 suggests increased spatial representation of threatened species.  The slight increase in mean score for scenario 2 may indicate a small increase in spatial representation of low-risk species, a small decrease in spatial representation of high-risk species, or a combination of both.  The large decrease shown in scenario 3 indicates that a zero threshold greatly increases the spatial representation of high-risk species.

Given that only `r format(p_both_all, big.mark = ',')`% of species overlap, it is surprising that changing the priority for overlapping species from IUCN maps to AquaMaps would result in such large country-level score shifts as seen in scenarios 2 and 3.  While many country scores showed only small variation across all three scenarios, select countries gained up to 5 points while others dropped as many as 12 points.  Significantly, the global score decreased by two full points in scenario 3, indicating that an arbitrary change in how the two datasets are combined can result in a more dire assessment of species conservation status.

@@

Klein et al. (2015) compare the global distribution of species to the global distribution of marine protected areas to assess how well the MPAs protect key species and identify which species fall through gaps in protection.  The study relied on the AquaMaps database, using a probability of occurrence threshold of 50% or greater, to determine species presence, and the World Database of Protected Areas to define zones of marine protection.  They found that the global MPA network leaves 90.5% of marine species with less than 5% of their overall range represented within MPAs, and 1.4% of species have no protection at all ("gap" species).

We recalculated the amount of under-protected and gap species using either IUCN or AquaMaps data (using 2015 AquaMaps data and a 0% threshold to allow the most meaningful comparison to IUCN's "limits of distribution", <a href = '#fig5'>Fig. 5</a>).  We found a five-fold increase in the proportion of gap species (6.4% of species vs. 1.2%) and dramatically larger proportion of species with less than 2% of their range protected (73.2% of species vs. 47.7%).  However, this comparison also indicates a larger proportion of well-protected species with greater than 10% of range protected (2.9% of species vs. 1.5%).  Conclusions drawn from each of these datasets would paint dramatically different pictures of the protection afforded by our current global MPA network.

@@

AquaMaps and IUCN range maps show reasonable agreement for many well-studied species, but substantial differences illustrate uncertainty in our understanding of spatial distribution for many others.  Although many other appriaches exist for species distribution modeling, these two are the only ones applied globally to marine species.  Identifying and addressing differences in these datasets will increase their utility for research and conservation actions.  Several likely drivers of commission errors and omission errors between these datasets point to a few important ways to improve range alignment.

For IUCN range data, clipping ranges to known depth limits improves output for many species, most notably corals and other demersal species dependent on photosynthesis.  If species' depth limits are not known, simple rules of thumb will likely reduce commission errors without introducing substantial omission errors.  For example, for most corals, researchers could clip range maps to the photosynthetic limit of 200 meters.  For AquaMaps range data, dependent primarily on environmental and physical preferences and conditions, implementing area restrictions based on biogeographical criteria such as Marine Ecoregions of the World (Spalding et al. 2007) would likely decrease commission errors and improve predictive power, especially for data-poor species.

Many studies using AquaMaps data test the sensitivity of results by varying the probability of occurrence threshold to determine presence.  This decision ultimately represents a tradeoff between errors of commission (low threshold) and omission (high threshold).  Using AquaMaps in conjunction with IUCN can mitigate potential errors, while also increasing the taxonomic and spatial breadth of coverage, as long as the differences between the datasets can be reasonably minimized. In this case, we recommend a presence threshold of 0% as it most closely approximates the "limits of distribution" criterion defined by IUCN data providers.

<!-- summarize back to big picture: CBD, MPAs, climate velocity, etc... understanding species distributions, and therefore minimizing errors, is important to all these things -->
_AquaMaps and IUCN are two of the most important databases providing species distribution information across a wide range of marine species, taxa, and regions.  The different methodologies behind the creation of each database is likely to introduce different rates of commission and omission errors, which can have a profound impact on our understanding of species ranges.  By comparing overlapping species maps from these two databases, we have identified likely sources of error, allowing data users to address or account for them.  For users of AquaMaps, IUCN, and other datasets of species distribution, reducing the likelihood of these errors is critical to effective biodiversity research and conservation action._

Human impacts on oceans (pollution, ocean acidification, overfishing, extractive use) are increasingly pressuring the marine ecosystems that provide food, economic opportunity, cultural value, and life support.  Climate change and direct human impacts are driving shifts in species richness and abundance.  Managing and protecting the oceans to deliver these benefits sustainably requires much deeper understanding of the flora and fauna.  Managing and protecting marine species improves the resilience of critical ocean ecosystems to human-induced pressures. But effective management depends critically on accurate and robust understanding of the distribution of species we want to protect.

### 2016-03-21 ms_3_methods.Rmd
Ben's comments, some fixes
incorporated Ben's comments; reframing a conclusion; preparing figures.  Moved the MPA analysis explanation of shifts into the SI

To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; species presence within a grid cell was determined by any non-zero overlap of a species polygon with the cell, and number of species per cell was simply the count of the species present.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.  We represented relative distribution of species count for each dataset by plotting average species count against latitude and longitude (Fig. A, 2B).

@@

**Methods for OHI case study: **
Using methods and supplemental materials from OHI (Halpern et al. 2012, 2015), we modified the original code for OHI 2015 Species status (SPP) (REF), allowing for flexibility in prioritized data source and AquaMaps presence threshold.  We ran the SPP code three times, prioritizing IUCN over AquaMaps for a 0% threshold, and prioritizing AquaMaps over IUCN for both a 40% and 0% threshold.  We compared each of these to the output of the published OHI 2015 SPP model (prioritizing IUCN over AquaMaps at a 40% threshold).

**Methods for MPA Gap Analysis case study:** 
Based upon the methods described in Klein et al. 2015, we reconstructed the analysis using the subset of protected areas (WDPA 2014) spatially covering a marine area and classified as IUCN I-IV. The WDPA polygons and marine polygons were rasterized to 0.01° and then aggregated to AquaMaps native 0.5° cells, to calculate proportion of protected area and marine within each cell.  After verifying our results using the 2014 AquaMaps dataset, we updated the analysis using the 2015 AquaMaps dataset at a presence threshold of 50% (to compare to Klein et al. directly) and 0% (to better compare with IUCN spatial data).  To analyze MPA coverage against IUCN spatial data, we extracted IUCN polygon weights per 0.5° cell for each species and compared against the protected area raster.  Finally, we combined AquaMaps data (at 0% threshold) and IUCN data, using AquaMaps for the `r n_spp_both` overlapping species and again compared against the protected area raster.

All processing was completed in R (REF), and all code and intermediate data are available on GitHub (REF?) at https://github.com/OHI-Science/IUCN-AquaMaps.

>>>>

To compare the spatial representation of the two datasets directly, we rasterized the IUCN species polygons to the same 0.5° grid as the AquaMaps species maps; species presence within a grid cell was determined by any non-zero overlap of a species polygon with the cell, and number of species per cell was simply the count of the species present.  For the AquaMaps dataset, we determined per-cell species count by including all species with non-zero probability of occurrence, to best approximate the "extent of occurrence" generally indicated by IUCN maps.  We represented relative distribution of species count for each dataset by plotting average species count against latitude and longitude (Fig. 2).

@@

**Methods for OHI case study: **
Using methods and supplemental materials from OHI (Halpern et al. 2012, 2015), we modified the original code for OHI 2015 Species status (SPP) (REF), allowing for flexibility in prioritized data source and AquaMaps presence threshold.  We ran the SPP code three times, prioritizing IUCN over AquaMaps for a 0% threshold, and prioritizing AquaMaps over IUCN for both a 40% and 0% threshold.  We compared each of these to the output of the published OHI 2015 SPP model (which prioritizes IUCN over AquaMaps at a 40% threshold).

**Methods for MPA Gap Analysis case study:** 
Based upon the methods described in Klein et al. 2015, we reconstructed the analysis using the subset of protected areas (WDPA 2014) spatially covering a marine area and classified as IUCN I-IV. The WDPA polygons and marine polygons were rasterized to 0.01° and then aggregated to AquaMaps native 0.5° cells, to calculate proportion of protected area and marine within each cell.  After verifying our results using the 2014 AquaMaps dataset, we updated the analysis using the 2015 AquaMaps dataset at a presence threshold of 50% (to compare to Klein et al. directly) and 0% (to better compare with IUCN spatial data).  To analyze MPA coverage against IUCN spatial data, we extracted IUCN polygon weights per 0.5° cell for each species and compared against the protected area raster.  Finally, we combined AquaMaps data (at 0% threshold) and IUCN data, using AquaMaps for the `r format(n_spp_both, big.mark = ',')` overlapping species and again compared against the protected area raster.

All processing was completed using R statistical software (REF), and all code and intermediate data are available on GitHub (REF?) at https://github.com/OHI-Science/IUCN-AquaMaps.
